---
title: "Project A"
author: "Ryan Dastrup"
date: "January 31, 2021"
output: html_notebook
---
### Description

This report will examine gun deaths and will examine the reasons of death, gun 
death by gender, and gun death by race. The data was housed on the website
Fivethirtyeight, and comes primarily from the Centers for Disease Control and
Prevention's Multiple cause of Death database. The years we will examine are
2012 and 2019. 


***

### Analysis

First we will examine the proportions of gun death in 2012. The biggest reason
for gun related causes was associated with suicide and (.62).  
```{r}

guns<-read.csv("full_data.csv")

```


```{r}

library(tidyverse)
library(knitr)
library(kableExtra)
library(ggplot2)
library(magrittr)

```


```{r}

death2012<-subset(guns, guns$year == "2012")

```


```{r}

More_death2012<- death2012 %>%
  group_by(intent) %>%
  summarize (freq = n()) %>%
  mutate (prop = freq/sum(freq))
death2012 %>%
  group_by(intent) %>%
  summarize(freq = n()) %>%
  mutate (prop = freq/sum(freq))

```


```{r}

ggplot(More_death2012, aes (x = intent, y = prop)) +
  geom_col(fill = "dodgerblue") +
  labs (x = "",
        y = "Proportion",
        title = "Reasons for Death - 2012")

```


```{r}

More_death2012 %>%
  knitr::kable(caption = "Proportions of Gun Death - 2012", 
               digits = 4) %>%
  kableExtra::kable_styling(bootstrap_options = "striped")

```



```{r include=FALSE}

More_gender2012<- death2012 %>%
  group_by(sex) %>%
  summarize (freq = n()) %>%
  mutate (prop = freq/sum(freq))
death2012 %>%
  group_by(sex) %>%
  summarize(freq = n()) %>%
  mutate (prop = freq/sum(freq))

```


```{r}

ggplot(More_gender2012, aes (x = sex, y = prop)) +
  geom_col(fill = "dodgerblue") +
  labs (x = "",
        y = "Proportion",
        title = "Gun Death by Gender - 2012")

```
```{r echo=FALSE}

More_gender2012 %>%
  knitr::kable(caption = "Table 2. Proportions for Gun Death by Gender in 2012", digits = 4) %>%
  kableExtra::kable_styling(bootstrap_options = "striped")

```


We had more deaths by men versus women. Males counted for 86% while women made up the rest (14%). 

```{r include=FALSE}

More_race2012<- death2012 %>%
  group_by(race) %>%
  summarize (freq = n()) %>%
  mutate (prop = freq/sum(freq))
death2012 %>%
  group_by(race) %>%
  summarize(freq = n()) %>%
  mutate (prop = freq/sum(freq))

```

```{r}

ggplot(More_race2012, aes (x = race, y = prop)) +
  geom_col(fill = "dodgerblue") +
  labs (x = "",
        y = "Proportion",
        title = "Gun Death by Race - 2012")

```


```{r echo=FALSE}

More_race2012 %>%
  knitr::kable(caption = "Gun Death by Race - 2012", digits = 5) %>%
  kableExtra::kable_styling(bootstrap_options = "striped")

```


Finally, we wanted to look at race related causes. We had five (5) different 
racial classes. We see gun death among white was more than half at (.65). Among
the black population we see about (.25) related deaths, with the remaining 
balance between Asian, Native American and Hispanics.

***

### Conclusion
To recap the biggest reason for gun death was due to suicide and we see that
majority of time this occurs with males. Further, the white population is by far
ahead of any of the other races examined.  


***

Next we will look at gun data for 2019, and we will examine the same variables 
as 2012.

```{r}

death_2019<- read.csv("full_data_19.csv")

```

```{r include=FALSE}

More_death_2019<- death_2019 %>%
  group_by(intent) %>%
  summarize (freq = n()) %>%
  mutate (prop = freq/sum(freq))
death_2019 %>%
  group_by(intent) %>%
  summarize(freq = n()) %>%
  mutate (prop = freq/sum(freq))

```

```{r echo=FALSE}

ggplot(More_death_2019, aes (x = intent, y = prop)) +
  geom_col(fill = "dodgerblue") +
  labs (x = "",
        y = "Proportion",
        title = "Reasons for Death - 2019")

```


```{r}

More_death_2019 %>%
  knitr::kable(caption = "Gun Death - 2019", digits = 4) %>%
  kableExtra::kable_styling(bootstrap_options = "striped")

```

```{r include=FALSE}

More_death_gender2019<- death_2019 %>%
  group_by(sex) %>%
  summarize (freq = n()) %>%
  mutate (prop = freq/sum(freq))
death_2019 %>%
  group_by(sex) %>%
  summarize(freq = n()) %>%
  mutate (prop = freq/sum(freq))

```

```{r echo=FALSE}

ggplot(More_death_gender2019, aes (x = sex, y = prop)) +
  geom_col(fill = "dodgerblue") +
  labs (x = "",
        y = "Proportion",
        title = "Gun Death by Gender - 2019")
```



```{r include=FALSE}

More_death_race2019<- death_2019 %>%
  group_by(race) %>%
  summarize (freq = n()) %>%
  mutate (prop = freq/sum(freq))
death_2019 %>%
  group_by(race) %>%
  summarize(freq = n()) %>%
  mutate (prop = freq/sum(freq))

```


```{r echo=FALSE}

ggplot(More_death_race2019, aes (x = race, y = prop)) +
  geom_col(fill = "dodgerblue") +
  labs (x = "Gun Death by Race - 2019",
        y = "Proportion",
        title = "Gun Death by Race - 2019")

```

```{r echo=FALSE}

More_death_race2019 %>%
  knitr::kable(caption = "Gun Death by Race - 2019", digits = 4) %>%
  kableExtra::kable_styling(bootstrap_options = "striped")

```


```{r}

data_12_19<- rbind(death2012, death_2019)
#data_12_19<-as_tibble(data_12_19)

```


```{r}

data_12_19_counts <- data_12_19 %>%
  group_by(year, intent) %>%
  summarize (freq = n(), .groups = 'drop')

data_12_19_IntentProps <- data_12_19_counts %>%
  group_by(intent) %>%
  mutate (prop = freq / sum(freq)) %>%
  ungroup()

```


```{r}

ggplot(data_12_19_IntentProps, aes (x = year, y = prop, fill = intent )) + 
  geom_col(state= "identity") +
  
  labs(x = '2012 and 2019', y = 'Proportion', title = "Reason for Death - 2012 and 2019") +
  scale_fill_discrete (name = 'intent',
                       labels = c ('Suicide', 'Homocide', 'Accidental', 'Undetermined'))
```


```{r echo=FALSE}

data_12_19_counts <- data_12_19 %>%
  group_by(year, intent) %>%
  summarize (freq = n(), .groups = 'drop')

data_12_19_IntentProps <- data_12_19_counts %>%
  group_by(intent) %>%
  mutate (prop = freq / sum(freq)) %>%
  ungroup()
data_12_19_IntentProps

```


```{r include=FALSE}

data_12_19racecounts <- data_12_19 %>%
  group_by(year, race) %>%
  summarize (freq = n(), .groups = 'drop')

data_12_19raceProps <- data_12_19racecounts %>%
  group_by(race) %>%
  mutate (prop = freq / sum(freq)) %>%
  ungroup()

data_12_19racecounts %>%
  knitr::kable(caption = "Gun Death by Race - 2019") %>%
  kableExtra::kable_styling(bootstrap_options = "striped")

```


```{r}
ggplot(data_12_19raceProps, aes (x = year, y = prop, fill = race )) +
  geom_col()+
  labs(x = '2012 and 2019', y = 'Proportion', title = "Gun Death by Race - 2012 and 2019") +
  scale_fill_discrete (name = 'race',
                       labels = c ('White', 'Black', 'Asian', 'Hispanic', 'Native American'))

```



```{r}

#data_12_19_Props 

data_12_19racecounts <- data_12_19 %>%
  group_by(year, race) %>%
  summarize (freq = n(), .groups = 'drop')

data_12_19raceProps <- data_12_19racecounts %>%
  group_by(race) %>%
  mutate (prop = freq / sum(freq)) %>%
  ungroup()

data_12_19raceProps

```


```{r include=FALSE}

data_12_19_counts <- data_12_19 %>%
  group_by(year, sex) %>%
  summarize (freq = n(), .groups = 'drop')

data_12_19_Props <- data_12_19_counts %>%
  group_by(sex) %>%
  mutate (prop = freq / sum(freq)) %>%
  ungroup()

```


```{r}

ggplot(data_12_19_Props, aes (x = year, y = prop, fill = 'sex')) +
  geom_col()+
  labs(x = '2012 and 2019', y = 'Proportion', title = "Gun Death by Gender - 2012 and 2019") +
  scale_fill_discrete (name = 'sex',
                       labels = c ('Female', 'Male'))

```


```{r}

data_12_19_Props 

```





